`Total-evidence' Bayesian estimation of phylogeny, divergence times and fossil ages

### 2016-04-28 response-to-reviewers.tex

I think the insistence by the authors to use the term "morhpological clock" will cause controversy. Please provide some evidence that departure of the clock is small (for example, posterior estimates of the rate variation parameter are, say, close to zero). Otherwise I suggest removing the term.

As a side note I would like to point out that there is an extensive literature on morphological rates of evolution, indeed, before molecular data became available in the sixties, most work on evolution was effectively work on phenotypic (morphological) evolution. One of the reasons the molecular clock was so controversial when it was initially proposed, was precisely because early workers were skeptical that molecular rates should be constant when morphological rates were so variable. For example, in his 1985 book on the neutral theory, Kimura has a whole chapter contrasting morphological vs. molecular rates (I strongly encourage the authors to read this), and he provides lots of citation to early works (for example, Haldane's) on morphological rates. Michael Lynch has some theoretical work where he shows that under neutral mutation and random drift, morphological clocks can be achieved (can't remember the reference) and he gives examples in humans. It may be worth looking this up if the authors want to add a balance discussion of this in their ms. There are also lots of papers throughout the 80s and 90s on morphological rates. One thing that surprises me with the new morphological papers is that a lot of this literature has been completely ignored. I encourage the authors to read and cite some of these works.

\response{We have indeed read some of this early literature and we thank the reviewer's encouragement to reference this earlier work on rates of phenotypic evolution. We have added a section entitled ``Rates of evolution and relaxed morphological clocks'':}

\respquote{In order to use discrete morphological comparative data to estimate fossil ages, it is necessary to assume a (relaxed) morphological clock. There is a long history of the study of the evolutionary rates of phenotypic characters \cite{Simpson1944,Haldane1949,Simpson1953,Gingerich1983,Gingerich1993}, going at least back to Darwin's Origin of Species \cite{Darwin1859}. Darwin noted that {\it ``Species of different genera and classes have not changed at the same rate''} and illustrated this point with examples of ``living fossils'' such as the Silurian mollusc {\it Lingula} \cite{Darwin1859}. However in the same chapter Darwin goes on to say {\it ``In members of the same class the average amount of change, during long and equal periods of time, may, perhaps, be nearly the same''}. Nevertheless, phenotypic evolution has more typically been characterized as not evolving in a clock-like manner, especially when compared to molecular evolution \cite{Kimura1983}. While there are many examples of extremely slow and fast rates of phenotypic evolution in the literature, we would argue that this is also true for molecular rates. We are not aware of a comprehensive and systematic comparison of variation in evolutionary rates at the phenotypic and molecular levels. Regardless, for the data sets that we analyze, we adopt the point of view that variation in the rate of phenotypic evolution across the phylogeny can be accommodated with a relaxed morphological clock.
}

### 2016-04-28 fossilDating-revision.tex
neutral evolution url
canid log files and morphological clock intro

We extend this approach into the realm of discrete morphological evolution by presenting a statistical model of evolution that generates an expectation on the distribution of fossils, their morphological characters. 
This model has been previously presented in the context of divergence time dating \cite{gavr2014,gavryushkina2015bayesian}. 
It is distinct from alternative divergence time dating approaches in that it provides an explicit treatment of the temporal information contained in fossil remains, whether or not related molecular sequence data is available. 
This leads to an estimate of the age of the most recent common ancestor of a group of fossil and extant taxa. A key difference between this approach and earlier approaches to tip-calibrated ``total-evidence'' dating \cite{Ronq2012} is the admission of a probability that each fossil taxon may represent a sampled ancestor of one or more taxa  in the tree \cite{gavr2014}.
We exploit this framework to attempt the estimation of the age of individual fossils based solely on morphological data and their phylogenetic affinities to related taxa of known age. 
The method is applied to two rich and well-characterized morphological data sets: (i) 
%TS how many extant penguins?
19 extant penguins and 36 fossil relatives \cite{ksepka2010,ksepka2012}, (ii) a sample of nine extant canids and \ncanidfossils{} fossil relatives \cite{Slater2015}.

@@

We applied two phylogenetic models to the penguin data set: \Mstrict{} and \Mrelaxed{}. \Mstrict{} assumed a strict morphological clock ($\mu$) and no gamma-distributed rate heterogeneity among sites. Model \Mrelaxed{} partitioned the alignment into six partitions, with the $i$'th partition containing all characters that had $k=i+1$ character states across the sampled taxa. Model \Mrelaxed{} also uses a uncorrelated lognormally-distributed relaxed molecular clock \cite{drummond2006} with parameters $\mu$ and $S$ for the mean rate and log standard deviation of the rates. We followed \cite{gavryushkina2015bayesian} in having a broad LogNormal(-5.5, 2) prior on $\mu$ for all analyses and a Gamma($\alpha=0.5396$, $\beta=0.3819$) prior on $S$ for the relaxed clock analyses.

@@

\subsection*{Penguins conform well to a strict morphological clock}
Although \Mstrict{} is a very simple model, the phylogenetic estimates of the ages of the penguin fossils were remarkably consistent with their palaeontological age ranges. 
Figure \ref{fig:phyloAgeVsGeoAge}a plots the geological age and range against the phylogenetic estimates of fossil age. The points in this plot have $R^2 = 0.903$. 
The median error (difference between the phylogenetic median and the geological median) is 1.96 Myr. A summary of the individual estimates are tabulated in Table \ref{fossilTable1}.

@@

In this paper we have demonstrated that even a small number of morphological characters (some of the fossils had as few as 7 morphological traits coded) can be used in the context of a rich fossil reference data set, to provide an accurate age of the fossil based on a phylogenetic model. 
We found that although a strict morphological clock does a surprisingly good job of estimating fossil ages, there is evidence that phylogenetic estimation of fossil ages is improved by a model that includes a relaxed morphological clock. 
We found that fossil estimates were both slightly more accurate and more precise on average in the canid data set. 
This may suggest that the larger reference set of fossils in the canid data set (115 versus 35) more than makes up for the smaller number of characters (122 versus 245) with regards to accuracy and precision of fossil age estimates. However since the average age of the canid fossils is considerably less than that for the penguin fossils, it could simply be that the absolute errors are smaller because the true values are smaller.

>>>>

We extend this approach into the realm of discrete morphological evolution by presenting a statistical model of evolution that generates an expectation on the distribution of fossils, their morphological characters. 
This model has been previously presented in the context of divergence time dating \cite{gavr2014,gavryushkina2015bayesian}.

It is distinct from alternative divergence time dating approaches in that it provides an explicit treatment of the temporal information contained in fossil remains, whether or not related molecular sequence data is available. 
This leads to an estimate of the age of the most recent common ancestor of a group of fossil and extant taxa. A key difference between this approach and earlier approaches to tip-calibrated ``total-evidence'' dating \cite{Ronq2012} is the admission of a probability that each fossil taxon may represent a sampled ancestor of one or more taxa  in the tree \cite{gavr2014}.
We exploit this framework to attempt the estimation of the age of individual fossils based solely on morphological data and their phylogenetic affinities to related taxa of known age. 
The method is applied to two rich and well-characterized morphological data sets: (i) 
%TS how many extant penguins?
19 extant penguins and 36 fossil relatives \cite{ksepka2010,ksepka2012}, (ii) a sample of nine extant canids and \ncanidfossils{} fossil relatives \cite{Slater2015}.

%AJD revision
\subsection*{Rates of evolution and relaxed morphological clocks}

In order to use discrete morphological comparative data to estimate fossil ages, it is necessary to assume a (relaxed) morphological clock. There is a long history of the study of the evolutionary rates of phenotypic characters \cite{Simpson1944,Haldane1949,Simpson1953,Gingerich1983,Gingerich1993}, going at least back to Darwin's Origin of Species \cite{Darwin1859}. Darwin noted that {\it ``Species of different genera and classes have not changed at the same rate''} and illustrated this point with examples of ``living fossils'' such as the Silurian mollusc {\it Lingula} \cite{Darwin1859}. However in the same chapter Darwin goes on to say {\it ``In members of the same class the average amount of change, during long and equal periods of time, may, perhaps, be nearly the same''}. Nevertheless, phenotypic evolution has more typically been characterized as not evolving in a clock-like manner, especially when compared to molecular evolution \cite{Kimura1983}. While there are many examples of extremely slow and fast rates of phenotypic evolution in the literature, we would argue that this is also true for molecular rates. We are not aware of a comprehensive and systematic comparison of variation in evolutionary rates at the phenotypic and molecular levels. Regardless, for the data sets that we analyze, we adopt the point of view that variation in the rate of phenotypic evolution across the phylogeny can be accommodated with a relaxed morphological clock.

%http://earth.geology.yale.edu/~ajs/1993/11.1993.17Gingerich.pdf
%https://books.google.co.nz/books?id=olIoSumPevYC&printsec=frontcover#v=onepage&q&f=false

%%%

@@

We applied two phylogenetic models to the penguin data set: \Mstrict{} and \Mrelaxed{}. \Mstrict{} assumed a strict morphological clock ($\mu$) and no gamma-distributed rate heterogeneity among sites. Model \Mrelaxed{} partitioned the alignment into six partitions, with the $i$'th partition containing all characters that had $k=i+1$ character states across the sampled taxa. Model \Mrelaxed{} also uses a uncorrelated lognormally-distributed relaxed molecular clock \cite{Drummond2006} with parameters $\mu$ and $S$ for the mean rate and log standard deviation of the rates. We followed \cite{gavryushkina2015bayesian} in having a broad LogNormal(-5.5, 2) prior on $\mu$ for all analyses and a Gamma($\alpha=0.5396$, $\beta=0.3819$) prior on $S$ for the relaxed clock analyses.

@@

\subsection*{Penguins conform well to a morphological clock}
Although \Mstrict{} is a very simple model, the phylogenetic estimates of the ages of the penguin fossils were remarkably consistent with their palaeontological age ranges. 
Figure \ref{fig:phyloAgeVsGeoAge}a plots the geological age and range against the phylogenetic estimates of fossil age. The points in this plot have $R^2 = 0.903$. 
The median error (difference between the phylogenetic median and the geological median) is 1.96 Myr. A summary of the individual estimates are tabulated in Table \ref{fossilTable1}.

@@

In this paper we have demonstrated that even a small number of morphological characters (some of the fossils had as few as 7 morphological traits coded) can be used in the context of a rich fossil reference data set, to provide an accurate age of the fossil based on a phylogenetic model.

%AJD
We found that although a strict morphological clock does a surprisingly good job of estimating fossil ages, there is evidence that phylogenetic estimation of penguin fossil ages is improved by a model that includes a variation in rates of morphological evolution among lineages. 
However, in the penguin data set the variation in evolutionary rates was not extreme and the log standard deviation of the relaxed morphological clock was comparable to values obtained for molecular clocks.   
%%%

We found that fossil estimates were both slightly more accurate and more precise on average in the canid data set. 
This may suggest that the larger reference set of fossils in the canid data set (115 versus 35) more than makes up for the smaller number of characters (122 versus 245) with regards to accuracy and precision of fossil age estimates. However since the average age of the canid fossils is considerably less than that for the penguin fossils, it could simply be that the absolute errors are smaller because the true values are smaller.
